\section{Simple correspondence analysis}\label{sec:ca-simple}
\ixon{correspondence analysis!two-way tables}
\subsection{Notation and terminology}\label{sec:ca-notation}
Because \CA\ grew up in so many homes, the notation, formulae
and terms used to describe the method vary considerably.
The notation used here generally follows \citet{Greenacre:84,Greenacre:97},
as does the documentation in the \STUGref{19}{The CORRESP Procedure}.

The descriptions here employ the following matrix and vector definitions:
\begin{itemize}
\item $\mat{N} = \{ n_{ij} \}$ is the $I \times J$ contingency table
with row and column totals $n_{i+}$ and $n_{+j}$, respectively.
The grand total $n_{++}$ is also denoted by $n$ for simplicity.
\item $\mat{P} = \{ p_{ij} \} = \mat{N}/n$ is the matrix of joint cell
probabilities,  called the \glossterm{correspondence matrix}.
\item $\vec{r} = \sum_j p_{ij} = \mat{P} \vec{1}$ is the row margin of $\mat{P}$;
$\vec{c} = \sum_i p_{ij} = \mat{P}\trans \vec{1}$ is the column margin.
$\vec{r}$ and $\vec{c}$ are called the \emph{row masses} and \emph{column masses}.
\item $\mat{D}_r$ and $\mat{D}_c$ are diagonal matrices with $\vec{r}$
and $\vec{c}$ on their diagonals, used as weights.
\item $\mat{R} = \mat{D}_r^{-1} \mat{P} = \{ n_{ij} / n_{+j} \}$ is the matrix of
row conditional probabilities, called \emph{row profiles}.
Similarly, $\mat{C} = \mat{D}_c^{-1} \mat{P}\trans = \{ n_{ij} / n_{i+} \}$ is the matrix of
column conditional probabilities or \emph{column profiles}.
\end{itemize}

Two types of coordinates, $X$, $Y$ for the row and column categories are defined,
based on the generalized singular value decomposition of $\mat{P}$,
\ix{singular value decomposition}
\begin{equation*}%\label{eq:ca-svd}
\mat{P} = \mat{A} \mat{D}_{\lambda} \mat{B}\trans
\end{equation*}
where $\mat{D}_{\lambda}$ is the diagonal matrix of singular values
\(\lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_M\),
$\mat{A}$ is the $I \times M$ matrix of left singular vectors,
normalized so that
\( \mat{A} \mat{D}_r^{-1} \mat{A}\trans = \mat{I} \), and
$\mat{B}$ is the $J \times M$ matrix of right singular vectors,
normalized so that
\( \mat{B} \mat{D}_c^{-1} \mat{B}\trans = \mat{I} \).
Thus the columns of $\mat{A}$ and $\mat{B}$ are orthogonal in the weighted metrics
defined by the row and column margins, $\mat{D}_r^{-1}$ and $\mat{D}_c^{-1}$,
respectively.
\begin{description}
\ix{correspondence analysis!principal coordinates}
\ix{principal coordinates}
\item[principal coordinates]  The coordinates of the row ($\mat{F}$) and column ($\mat{G}$) profiles
with respect to their own principal axes are defined so that the inertia along
each axis is the corresponding singular value, $\lambda_i$,
\begin{eqnarray}
%
\mat{F} & = & \mat{D}_r^{-1} \mat{A} \mat{D}_{\lambda} \quad\mbox{so that} \quad \mat{F}\trans \mat{D}_r \mat{F} = \mat{D}_{\lambda} \label{eq:pcoord1} \\
\mat{G} & = & \mat{D}_c^{-1} \mat{B} \mat{D}_{\lambda} \quad\mbox{so that} \quad \mat{G}\trans \mat{D}_c \mat{G} = \mat{D}_{\lambda} \label{eq:pcoord2}
\end{eqnarray}
\ix{correspondence analysis!standard coordinates}
\ix{standard coordinates}
\item[standard coordinates] The standard coordinates ($\mat{\Phi}, \mat{\Gamma}$) are a rescaling of the principal
coordinates to unit inertia along each axis,
\begin{eqnarray}
%\label{}
\mat{\Phi} & = & \mat{D}_r^{-1} \mat{A}  \quad\mbox{so that} \quad \mat{\Phi}\trans \mat{D}_r \mat{\Phi} = \mat{I} \\
\mat{\Gamma} & = & \mat{D}_c^{-1} \mat{B} \quad\mbox{so that} \quad \mat{\Gamma}\trans \mat{D}_c \mat{\Gamma} = \mat{I}
\end{eqnarray}
These differ from the principal coordinates in \eqref{eq:pcoord1}
and \eqref{eq:pcoord2} simply by the absence of the scaling factors,
$\mat{D}_{\lambda}$.
\end{description}
Thus, the weighted average of the squared principal coordinates
for the rows or columns on a principal axis equals the squared
singular value, $\lambda$ for that axis,
whereas the weighted average of the squared standard coordinates
equals 1.
The relative positions of the row or column points along any axis
is the same under either scaling,
but the distances between points differ, because the axes are
weighted differentially in the two scalings.


\input{ch5/properties}
\subsection{The CORRESP Procedure}\label{sec:corresp}

Correspondence analysis is performed using \PROC{CORRESP} in \SSTAT.
\PROC{CORRESP} can read two kinds of input:

\begin{itemize*}
\item a two-way contingency table (\emph{contingency table form}),
where the columns are \Dset{}
variables (specified in a \stmt{var}{CORRESP}), and the rows are observations, labeled by an
\pname{id} variable.
In this case the column variables contain the frequencies in the
corresponding cells.
\ix{case form}
\ix{frequency form}
\item raw category responses (\emph{case form}), or cell frequencies (\emph{frequency
form}),
classified by two (or more) table variables.  In these two cases, the table variables are specified in
a \stmt{TABLES}{CORRESP}.  When the observations are cell frequencies,
the frequency variable may be specified in the \stmt{WEIGHT}{CORRESP}.
\end{itemize*}

In addition to printed output, the \opt{OUTC=}{CORRESP} produces an \ODS\
which contains the row and column
coordinates and other information.
To understand the relationships among the row and column categories
we may plot the coordinates with
\PROC{PLOT} or \PROC{GPLOT}.  The
procedure has many options for scaling row and column coordinates,
and for printing various statistics which aid interpretation.
We first illustrate the basic use of the procedure.
A macro program \pname{CORRESP} described in \secref{sec:ca-camacro}
simplifies the analysis and plotting steps.

\input{ch5/haireye}
\input{ch5/camacro}
\input{ch5/victims2}
\input{ch5/quasi}
\ixoff{correspondence analysis!two-way tables}
\input{ch5/scores}
