GLMs for binary outcomes
========================================================

Example: Passengers on the Titanic

----------------------------------

```{r setup, include=FALSE}
load("c:/sasuser/catdata/R/data/ptitanic.rda")

```

Data on `r nrow(ptitanic)` passengers on the Titanic is recorded
in the data.frame `ptitanic`.  We fit a logistic regression model
for `survived` using a generalized linear model,
with passenger class (`pclass`), `age`, and `sex` as predictors.


Other analyses have shown that there is at least an interaction
between passenger class and sex, so that term is included in
the basic model.  As in model formulas, `pclass*sex` is
a short-hand for `pclass + sex + pclass:sex`, including the
main effects as well.

```{r}
titanic.glm = glm(survived ~ pclass*sex + age,
  family = binomial(logit), data = ptitanic)
summary(titanic.glm)
```

We can plot the high-order terms in this model using the effects package:

```{r titanic-effplot1, fig.width=7, fig.height=4}
# effect plots
library(effects)
titanic.eff <- allEffects(titanic.glm)
# All high-order effects
plot(titanic.eff)

```

