## Chapter 3

### Exercise 3.1 

The Arbuthnot data in HistData (Example 3.1) also contains the variable Ratio,
giving the ratio of male to female births.

a. Make a plot of `Ratio` over `Year`, similar to Figure 3.1. What features stand out? 
Which plot do you prefer to display the tendency for more male births?
```{r ex3.1a}
library(HistData)
data(Arbuthnot, package ="HistData")

# plot of Ratio by Year
par(mar=c(5,4,1,1)+.1)
with(Arbuthnot, {
  plot(Year, Ratio, type='b', ylim=c(.95, 1.2), 
       ylab="Birth Ratio (Male/Female)")
  abline(h=1, col="green", lwd=1)
  abline(h=mean(Ratio), col="red")
  text(x=1660, y=1, "Equal M/F ratio", pos=1, col="green3") 
  Arb.smooth <- loess.smooth(Year,Ratio)
  lines(Arb.smooth$x, Arb.smooth$y, col="blue", lwd=2)
})
summary(Arbuthnot$Ratio)

```

The plot is similar to Figure 3.1 in the text. If it is easier to think in terms of probability of a male birth,
plotting that directly may be preferable.

b. Plot the total number of christenings, `Males + Females` or `Total` (in 000s) over time.
What unusual features do you see?

```{r ex3.1b}
# total number of Christenings
with(Arbuthnot, {
  Total= Males + Females
  plot(Year, Total, type='b', ylab="Total Christenings (Male + Female)")
  Arb.smooth <- loess.smooth(Year,Total)
  lines(Arb.smooth$x, Arb.smooth$y, col="blue", lwd=2)
})
```

There was a large decline in births between 1640--1660, corresponding to years of plague in England.

### Exercise 3.2

Use the graphical methods illustrated in Section 3.2 to plot a collection of geometric distributions for $p$ = 0.2, 0.4, 0.6, 0.8, over a range of values of $k$ = 0, 1, ... 10.

a. With `xyplot()`, try the different plot formats using points connected with lines, as in Figure 3., or using points and lines down to the origin, as in the panels of Figure 3.10.

```{r ex3.2.a}
KL <- expand.grid(k = 0 : 10, p = c(0.2, 0.4, 0.6, 0.8))
geom_df <- data.frame(KL, prob = dgeom(KL$k, KL$p))
geom_df$p = factor(geom_df$p)
str(geom_df)

library(lattice)
mycol<-palette()[2:5]
xyplot(prob ~ k | p , data = geom_df, type = c("b"), 
       pch = 16, lwd = 4, cex = 1.25, 
       xlab = list("Number of events (k)", cex = 1.25), layout = c(2,2), 
       ylab = list("Probability", cex = 1.25))
xyplot(prob ~ k | p , data = geom_df, type = c("h", "p"), 
       pch = 16, lwd = 4, cex = 1.25, 
       xlab = list("Number of events (k)", cex = 1.25), layout = c(2,2), 
       ylab = list("Probability", cex = 1.25))
```

b. Also with `xyplot()`, produce one version of a multi-line plot in a single panel that you think shows well how these distributions change with the probability $p$ of success.

```{r ex3.2.b}
geomplt<-xyplot(prob ~ k , data = geom_df, groups = p, 
                type = c("b"), pch = 16, lwd = 4, cex = 1.25, col = mycol, 
                xlab = list("Number of events (k)", cex = 1.25), 
                ylab = list("Probability", cex = 1.25))
library(directlabels)
direct.label(geomplt, list("top.points", cex = 1.5, dl.trans(y = y + 0.1)))
```

c. Do the same in a multi-panel version, conditional on $p$

```{r ex3.2.c}
xyplot(prob ~ k | p , data = geom_df, type = c("b"), 
       pch = 16, lwd = 4, cex = 1.25, 
       xlab = list("Number of events (k)", cex = 1.25), layout = c(4,1), 
       ylab = list("Probability", cex = 1.25))
```

### Exercise 3.3

Use the data set `WomenQueue` to:

a. Produce plots analogous to those shown in Section 3.1 (some sort of bar graph of frequencies)
```{r ex3.3a}
data("WomenQueue", package = "vcd")
barplot(WomenQueue,xlab="Number of Women in queues of 10",ylab= "Frequency")
```

b. Check for goodness-of-fit to the binomial distribution using the `goodfit()` methods described
in Section 3.3.2.

Note that with `goodfit()`, you should specify $n=10$ for the binomial distribution as the `size` parameter.
```{r ex3.3b}
library(vcd)
gf.women <- goodfit(WomenQueue, type = "binomial", par=list(size=10))
summary(gf.women)

```

c. Make a reasonable plot showing departure from the binomial distribution.

The simplest plot is the hanging rootogram.  An alternative plot is a "binomialness" plot produced by `distplot()`.
```{r ex3.3c}
plot(gf.women, xlab = "Queue Length")
distplot(WomenQueue, type = "binomial", size=10, xlab = "Queue Length")
```

d. Suggest some reasons why the number of women in queues of length 10 might depart from a binomial distribution, $Bin(n=10, p=1/2)$.

- Perhaps women (or men) are more prevalent in these queues, so $p \ne 1/2$.
- People often join lines in groups, so the observations are unlikely to be independent.

### Exercise 3.4

Continue Example 3.13 on the distribution of male children in families in Saxony by fitting a binomial distribution, 
$Bin(n = 12; p = 1/2 )$, specifying equal probability for boys and girls. 

a. Carry out the GOF test for this fixed binomial distribution. What is the ratio of Chi-sqrare/df? What
do you conclude?

Note that you need to specify both $n$ and $p$ as fixed parameters here.
```{r ex3.4a}
Saxony_gf <-goodfit(Saxony, type = "binomial", par=list(size=12, prob=.5))
ss <-summary(Saxony_gf)
# The ratio of Chi-square/df
ss[,"X^2"] / ss[,"df"]
```
The binomial model fits very badly.

b. Test the additional lack of fit for the model Bin(n = 12; p = 1/2 ) compared to the model
$Bin(n = 12; p = \hat{p})$ where $\hat{p} is estimated from the data.

```{r ex3.4b}
Saxony_gf2 <- goodfit(Saxony, type = "binomial", par=list(size=12))
summary(Saxony_gf2)
```
This fits much better, but still not a good fit. 

c. Use the `plot.goodfit()` method to visualize these two models.

```{r ex3.4c}
plot(Saxony_gf, main = "Fit for p=0.5", xlab = "Number of Male Children (Out of 12)")
plot(Saxony_gf2, main = "Fit for p=phat", xlab = "Number of Male Children (Out of 12)")
```

### Exercise 3.5

For the `Federalist` data, the examples in Section 3.3.1 and Section 3.3.2 showed the negative binomial to provide an acceptable fit. Compare this with the simpler special case of geometric distribution, corresponding to $n = 1$.

a. Use `goodfit()` to fit the geometric distribution. [Hint: use `type="nbinomial"`, but specify `size=1` as a parameter.]

```{r ex3.5a}
fdfit1 <- goodfit(Federalist, type = "binomial", par = list(size=6))
fdfit1
fdfit2 <- goodfit(Federalist, type = "nbinomial", par = list(size=1))
fdfit2
```

b. Compare the negative binomial and the geometric models statistically, by a likelihood-ratio test of the difference between these two models.

<!--
```{r ex3.5b}
library(MASS)
block<-as.numeric(names(Federalist))
wordc<-as.vector(Federalist)
fd.df<-data.frame(block, wordc)
fd.df
fd.nbin <- glm.nb(wordc ~ block, method = "glm.fit", data = fd.df) #Does not converge, iteration limit reached
fd.geom <- glm(cbind(1, wordc)~ block, family = binomial, data = fd.df)
LRstats(fd.nbin, fd.geom)
```
-->

c. Compare the negative binomial and the geometric models visually by hanging rootograms or other methods.

```{r ex3.5.c}
plot(fdfit1)
plot(fdfit2)
distplot(Federalist, type = "binomial", size=6, xlab = "Word Count")
distplot(Federalist, type = "nbinomial", size=6, xlab = "Word Count")
```

### Exercise 3.6 

Mosteller and Wallace (1963, Table 2.4) give the frequencies, $n_k$, of counts $k =
0, 1, \dots$ of other selected marker words in 247 blocks of text known to have been written by Alexander Hamilton. The data below show the occurrences of the word *upon*, that Hamilton used much
more than did James Madison.

a. Read these data into R and construct a one-way table of frequencies of counts or a matrix or
data frame with frequencies in the first column and the corresponding counts in the second
column, suitable for use with `goodfit()`.

`goodfit()` requires its first argument to be either a one-way table (from `xtabs()`),
or a data.frame with frequencies in the *first* column and the corresponding counts in the second column.  Both of the following forms will work.
```{r upon1}
count <- 0:5
Freq <- c(129, 83, 20, 9, 5, 1)
sum(Freq)  # check N

(Upon <- data.frame(Freq, count))             # as a data.frame
(Upon.tab <- xtabs(Freq ~ count, data=Upon))  # one-way table

```

b. Fit and plot the Poisson model for these frequencies.
```{r upon2}
(up0 <- goodfit(Upon, type="poisson"))
summary(up0)
plot(up0)
```

c. Fit and plot the negative binomial model for these frequencies.
```{r}
(up1 <- goodfit(Upon, type="nbinomial"))
summary(up1)
plot(up1)
```

d. What do you conclude?

### Exercise 3.7

### Exercise 3.8

### Exercise 3.9

### Exercise 3.10

### Exercise 3.11

### Exercise 3.12
