---
title: "DDAR: Solutions and Hints for Exercises"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: solutions-setup.docx
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo=FALSE}
library(vcdExtra)
options(digits=4)
knitr::opts_knit$set(global.par = TRUE)
par(mar=c(4,4,1,1)+.1, ps=8)
lattice::lattice.options(default.theme = list(fontsize = list(text = 10)))
```

This document is intended as an aid to instructors who wish to use
_Discrete Data Analysis with R_ in a course.  It contains the text
of the *Exercises* sections from all chapters, together with 
some solutions or hints for the various problems.

## Chapter 1: Introduction

These exercises are all conceptual. There are no hints or solutions.
The goal here was to get students to investigate various sources of
data graphics and think about how well they communicated some facts,
relationships or explanations.

```{r child="ex02.Rmd"}
```

```{r child="ex03.Rmd"}
```

```{r child="ex04.Rmd"}
```

```{r child="ex05.Rmd"}
```

```{r child="ex06.Rmd"}
```

```{r child="ex07.Rmd"}
```

