We are grateful to many colleagues, friends, students and Internet acquaintences who have contributed to
to this book, directly or indirectly.

We thank those who read and commented on various drafts of the book or chapters.
In particular, John Fox, Michael Greenacre, and several anonymous reviewers gave
insightful comments on the organization of the book and made many helpful suggestions.
Matthew Sigal used his wizardly skills to turn sketches of conceptual diagrams into
final figures. Phil Chalmers contributed greatly with technical and stylistic editing
of a number of chapters.

At a technical level, we were aided by the cooperation of a number of \R package authors, who
helped to enhance the graphic displays:
Achim Zeilleis who served as a guiding hand in the development of the \pkg{vcd} and \pkg{vcdExtra} packages;
John Fox and Sandy Weisberg for enhancements to the \pkg{car} and \pkg{effects} packages;
Milan Bouchet-Valat for incorporating suggestions dealing with plotting \func{rc} solutions into the \Rpackage{logmult};
Michael Greenacre and Oleg Nenadic for help to enhance plotting in the \Rpackage{ca};
Heather Turner for advice and help with plotting models fit using the \Rpackage{gnm};
Jay Emerson for improvements to the \Rpackage{gpairs}.

There were also many contributors from the R-Help email list (\url{r-help@r-project.org}), too many to name them all.
Special thanks for generous assistance go to:
David Carlson,
William Dunlap,
Bert Gunter,
Jim Lemon,
Duncan Murdoch,
Denis Murphy,
Jeff Newmiller,
Richard Heiberger,
Thierry Onkelinx,
Marc Schwartz,
David Winsemius, and
Ista Zahn.

The book was written using the \Rpackage{knitr}, allowing a relatively seamless 
integration of \LaTeX\  text, \R code, and \R output and graphs, so that any changes
in the code were automatically incorporated in the book.
Thanks are due to Yihui Xie and all the contributors to the \code{knitr} project
for making this possible.  We are also grateful to Phil Chalmers and Derek Harnanansingh
for assistance in using GitHub to manage our collaboration.
 Pere Mill\'an-Mart\'inez and Marcus Fontaine helped considerably with \LaTeX\
formatting issues.

The first author's work on this project was supported by grants from the
National Science and Engineering Research Council of Canada (Grant 8150)
and a sabbatical leave from York University in 2013--14, during which most
of this book was written.

