\chapter{Introduction}\label{ch:intro}
\input{ch1/vtoc}
\begin{quote}
{\Large
Categorical data consists of variables whose values comprise a set
of discrete categories.
Such data require different statistical and graphical methods
than commonly used for quantitative data.
The focus of this book is on visualization techniques and graphical
methods designed to reveal patterns of relationships among
categorical variables.
}
\end{quote}
\minitoc
\clearpage

\section{Data visualization and categorical data}
\input{ch1/beauty}
``Data visualization'' is an approach to data analysis which focuses
on \emph{insighful graphical display}.  We may display the raw data, some
summary statistics, or some indicators of the quality or adequacy
of a fitted model.  The word ``insightful'' suggests that the goal
is (hopefully) to reveal some aspects of the data which might not
be perceived, appreciated, or absorbed by other means.  The overall
aims include both beauty and truth, though each of these are 
only as perceived by the beholder.

Methods for visualizing quantitative data have a long history
and are now widely used in both data analysis and in data presentation, and
in both popular and scientific media.
Graphical methods for categorical data,
however, have only recently developed, and are consequently
not as widely used.  The goal of this book is to show concretely how
data visualization may be usefully applied to categorical data.

``Categorical data'' means different things in different
contexts.  We introduce the topic in \secref{sec:intro-whatis}
with some examples illustrating
\begin{seriate}
\item types of categorical variables: binary, nominal, and ordinal,
\item data in case form vs.\ frequency form,
\item frequency data vs.\ count data,
\item univariate, bivariate, and multivariate data, and
\item the distinction between explanatory and response variables.
\end{seriate}

Methods for the analysis of categorical data also fall into two
quite different categories, described and illustrated in \secref{sec:intro-strat}: the simple randomization-based
methods typified by
the classical Pearson $\chi^2$, Fisher's exact test, and Cochran-Mantel-Haenszel
tests, and the model-based methods represented by
logistic regression, loglinear, and generalized linear models.
\chrange{ch:discrete}{ch:corresp}
are mostly related to the randomization-based methods; 
\chrange{ch:logistic}{ch:loglin}
illustrate the model-based methods.

In \secref{sec:intro-grmeth} we describe some important similarities
and 
differences between categorical data and
quantitative data, and discuss the implications of these differences for
visualization techniques.
\secref{sec:intro-visualize} outlines a strategy of data analysis
focused on visualization.

\input{ch1/whatis}
\input{ch1/strategies}
%\section{Sampling models for categorical data}

\section{Graphical methods for categorical data}\label{sec:intro-grmeth}
\epigraph{You can see a lot, just by looking}{Yogi Berra}

The graphical methods for categorical data described in this book
are in some cases straightforward adaptations of more familiar
visualization techniques developed for quantitative data.
The graphical principles and strategies, and the relations between
the visualization approach and traditional statistical methods
are described in \SSSG, Chapter 1, and \cite{Cleveland:VisData}.
Another perspective on visual data display is presented in \secref{sec:intro-goals}.
However, the discrete nature of categorical data implies that
some familiar graphic methods need to be adapted, while in other
cases we require a new graphic metaphor for data display.
These issues are illustrated in \secref{sec:intro-catdata}.

\input{ch1/goals}
\input{ch1/catdata}
\input{ch1/visualize}
