\chapter{Two-way contingency tables}\label{ch:twoway}
\input{ch3/vtoc}
\begin{quote}
{\Large\noindent
The analysis of two-way frequency tables concerns the association
between two variables.  Different specialized displays are focused
on visualizing an odds ratio ($2 \times 2$ tables), or the general
pattern of association, or the agreement between row and column
categories.}
\end{quote}
\minitoc
\clearpage

\section{Introduction}
\epigraph{If you choose to represent the various parts in life by
holes upon a table, of different shapes---some circular, some
triangular, some square, some oblong---we shall generally find that
the triangular person has got into the square hole, the oblong into
the triangular, and a square person has squeezed himself into the
round hole.}{Sydney Smith, 1769-1845}

Most methods of statistical analysis are concerned with understanding
relationships among variables.
With categorical variables, these relationships are usually
studied from data which has been
summarized by a
\glossterm{contingency table},
giving the frequencies of observations cross-classified
by two or more such variables.

This chapter is concerned with methods for understanding the association
between two categorical variables.
Some simple examples are also presented which involve a third, stratifying variable,
where we wish to determine if the relationship between two primary
variables is the same or different for all levels of the
stratifying variable.
Methods for fitting models and displaying associations for three-way
and larger tables are described in \chref{ch:mosaic}.

We describe briefly some methods for testing whether an association
exists between two variables and for quantifying the strength
of this association (\secref{sec:twoway-tests}).
In \secref{sec:twoway-strat} we extend these ideas to situations where
the relation between two variables is of primary interest, but there
are one or more background variables to be controlled.

The main emphasis, however, is on graphical methods which help
to describe the nature of an association between variables.
\secref{sec:twoway-fourfold} presents the fourfold display,
designed to portray the odds ratio in $2 \times 2$ tables or a set
of $k$ such tables.
Sieve diagrams (\secref{sec:twoway-sieve}) and association plots
(\secref{sec:twoway-assoc}) are more general methods for depicting
the pattern of associations an any two-way tables.
When the row and column variables represent the classifications
of different raters, specialized measures and visual displays
for inter-rater agreement (\secref{sec:twoway-agree}) are particularly
useful.
Another specialized display, the trilinear plot (\secref{sec:twoway-trilinear})
is designed
for three-column frequency tables or compositional data.
In order to make clear some of the distinctions which occur in
\ctab{} analysis, I begin with several examples.

\begin{Example}[berkeley1]{Berkeley admissions}
\tabref{tab:berk22} shows aggregate data on applicants to
graduate school at Berkeley for the six largest departments in 1973
classified by admission and gender
\citep{Bickel-etal:75}.
See \datref{dat:berkeley} for the complete data set.
For such data we might wish to study whether there is an association
between admission and gender.
Are male (or female) applicants more likely to be admitted?
The presence of an association might be considered as
evidence of sex bias in admission practices.

\tabref{tab:berk22} is an example of the simplest kind of \ctab,
a $2 \times 2$ classification of individuals according to two
dichotomous (binary) variables.
For such a table, the question of whether there is an association
between admission and gender is equivalent to asking if the
proportions of males and females who are admitted to graduate
school are the same, or whether the difference in proportions
admitted is zero.
\input{ch3/tab/berk22}
\end{Example}

Although the methods for quantifying association in larger tables can be
used for $2 \times 2$ tables, there are specialized measures
(described in \secref{sec:twoway-tests}) and
graphical methods for these simpler tables.

It is frequently useful to make a distinction between \glossterm{outcome},
or response variables, on the one hand, and possible \glossterm{explanatory}
or predictor variables on the other.
In \tabref{tab:berk22}, it is natural to consider admission
as the outcome, and gender as the explanatory variable.
In other tables, no variable may be clearly identified as \emph{the}
outcome, or there may be several response variables.

\begin{Example}[haireye1]{Hair color and eye color}
\tabref{tab:hairdat} shows data collected by
\citet{Snee:74}
on the relation between hair color and eye color among 592
students in a statistics course
(\datref{dat:haireye}).  Neither hair color nor eye color
is considered a response in relation to the other;  our interest concerns
whether an association exists between them.
Hair color and eye color have both been classified
into four categories.  Although the categories used are among the most
common, they are not the only categories possible.%
\footnote{If students had been asked to write down their hair and eye
colors, it is likely that many more than four categories of each
would appear in a sample of nearly 600.}
Everyday observation suggests that there probably is an association
between hair color and eye color, and we will describe tests
and measures of associations for larger tables in
\secref{sec:twoway-overall}.
\input{ch3/tab/hairdat}
\end{Example}

Table variables may be treated simply as unordered,
\glossterm{nominal} variables,  or as
\glossterm{ordinal} variables, where the categories are
ordered according to some underlying dimension.
For example, the categories of hair color or eye color
are typically considered as nominal categories;
however, they
might arguably be considered to be ordered from light to dark.
When the variables are ordinal, more sensitive
and therefore more powerful tests of association can be used.

If, as we suspect, hair color and eye color are associated,
we would like to understand \emph{how} they are associated.
The graphical methods described later in this chapter help
reveal the pattern of associations present.

\begin{Example}[arthrit1]{Arthritis treatment}
The data in \tabref{tab:arthrit} compares an active treatment for rheumatoid
arthritis to a placebo
\citep{KochEdwards:88}.
The outcome reflects
whether individuals showed no improvement, some improvement, or
marked improvement.
Here, the outcome variable is an ordinal one, and it is probably
important to determine if the relation between treatment and outcome
is the same for males and females.
The data set is given more fully in \datref{dat:arthrit}.

This is, of course, a three-way table, with factors
Treatment, Sex, and Improvement.
If the relation between treatment and outcome is the same for
both genders, an analysis of the Treatment by Improvement
table (collapsed over sex) could be carried out.
Otherwise we could perform separate analyses for
men and women, or
treat the combinations of treatment and sex as four levels of
a ``population'' variable, giving a $4 \times 3$ two-way table.
These simplified approaches each ignore certain information
available in
an analysis of the full three-way table.
\input{ch3/tab/arthrit}
\end{Example}

\section{Tests of association for two-way tables}\label{sec:twoway-tests}

\subsection{Notation and terminology}\label{sec:twoway-notation}
To establish notation, let \(\mat{N}  =  \{  n_{ij}  \}\) be the
observed frequency table of variables \(A\) and \(B\) with \(r\) rows
and \(c\) columns, as shown in \tabref{tab:rbyc}.
In what follows, a subscript is replaced by a ``$+$''
when summed over the corresponding variable, so \(n_{i+}  =  \sum_j \,
n_{ij}\) gives the total frequency in row \(i\), \(n_{+j}  =  \sum_i \,
n_{ij}\) gives the total frequency in column \(j\), and \(n_{++}  =
\sum_i \sum_j \,  n_{ij}\) is the grand total; for convenience,
\(n_{++}\) is also symbolized by \(n\).
\input{ch3/tab/rbyc}

When each observation is randomly sampled from some population
and classified on two categorical variables, $A$ and $B$,
we refer to the \glossterm{joint distribution} of these variables,
and let $\pi_{ij} = \Pr(A=i,\,B=j)$ denote the probability that
an observation is classified in row $i$, column $j$ (or cell $(ij)$)
in the table.
Corresponding to these population joint probabilities, the
cell proportions, $p_{ij} = n_{ij} / n$, give the sample joint
distribution.

The row totals $n_{i+}$ and column totals $n_{+j}$ are called marginal frequencies for variables $A$ and $B$ respectively.
These describe the distribution of each variable \emph{ignoring} the other.
For the population probabilities, the \glossterm{marginal distributions}
are defined analogously as the row and column totals of the 
joint probabilities,
$\pi_{i+} = \sum_j \pi_{ij}$, and
$\pi_{+j} = \sum_i \pi_{ij}$.
The sample marginal proportions are, correspondingly, 
$p_{i+} = \sum_j p_{ij} = n_{i+} / n$, and 
$p_{+j} = \sum_i p_{ij} = n_{+j} / n$.

When one variable (the column variable, $B$, for example) is a response
variable, and the other ($A$) is an explanatory variable,
it is most often useful to examine the distribution of the response $B$
for \emph{each} level of $A$ separately.
These define the \glossterm{conditional distributions} of $B$, given the
level of $A$, and are defined for the population as
$\pi_{j\given i} = \pi_{ij} / \pi_{i+}$.

These definitions are illustrated in \outref{out:berkfreq.1}.
For the Berkeley data (\tabref{tab:berk22}), it shows the joint frequencies, $n_{ij}$, and joint sample percentages,
$100 \times p_{ij}$, in the first two rows within each table cell.
The third row in each cell (``Row pct'')
gives the conditional percentage of admission or rejection,
$100 \times p_{j\given i}$ for males and females separately.
The row and column labelled ``Total'' give the
marginal frequencies, $n_{i+}$ and $n_{+j}$,
and marginal percentages, $p_{i+}$ and $p_{+j}$.

\begin{Output}[htb]
\caption{Admission to Berkeley graduate programs: joint, marginal, and conditional percents}\label{out:berkfreq.1}
\small
\verbatiminput{ch\thechapter/out/berkfreq.1}
\end{Output}

\input{ch3/twobytwo}

\input{ch3/overall}
\input{ch3/stratified}

\input{ch3/fourfold}
%\input{ch3/tukey}
\input{ch3/sieve}

\input{ch3/assoc}
\input{ch3/agree}
\input{ch3/trilinear}

\input{ch3/summary}
%\section{Further reading}
